---
title: "Exploring the sentiment of trail users' tweets."
author: "Tom Wilson and Robin Lovelace"
date: "April 17, 2015"
output: ioslides_presentation
graphics: yes
---

## Presentation Overview

- Research Question
- Project Background
- Significance of Research
- Data
- Methods
- Selected Findings
- Known Issues
- Future Research

## Research Question

### Can the sentiment expressed in trail users' tweets help to assess the effectiveness of Environmental Stewardship Scheme (ESS) Agreements? A case study of the Pennine Way National Trail.

## Project background

```{r, echo=FALSE, fig.width=8, fig.align="left", fig.cap="The Pennine Way National Trail, England."}
library(png)
library(grid)
img <- readPNG("/Users/tom/Documents/pennine-tweets/PWNT_image01.png")
grid.raster(img)
```

## England's National Trails

- England has 15 National Trails which pass through diverse landscapes and expanses of agricultural land
- Original concept based upon the Appalachian Trail in the eastern US
- April 25th 2015 marks the 50th anniversary of England’s first National Trail - The Pennine Way (PWNT) 
- Annually 12 million visits made to England’s National Trails (Ramblers, 2012). 
- 300,000 average annual visitors to the PWNT (based upon Natural England trail counter data)

## The Pennine Way National Trail

```{r, echo=FALSE, fig.width=4, fig.height=4, fig.align="center", fig.cap="Location of the Pennine Way National Trail, England."}
library(png)
library(grid)
img <- readPNG("/Users/tom/Pictures/PennineWayEngland.png")
grid.raster(img)
```

- The Pennine Way National Trail travels 431 km (268 miles) along the central upland spine of England between Edale, Derybyshire to Kirk Yelthom in the Scottish Borders

## Environmental Stewardship Scheme (ESS)
- Provides government-financed payments to farmers or land managers in return for an environmentally sensitive approach to farming
- Most widespread approach to environmental management in England with agreements in place on over 70% of agricultural land
- 74.08% within 5km of Pennine Way
- Farmers can select from 65 management options which contribute toward a per-hectare total
- Meet point thresholds to secure a bi-annual payment.

## ESS Agreements within 5 and 25km of PWNT

```{r, echo=FALSE, fig.width=8, fig.height=4, fig.align="center", fig.cap="Location of the Pennine Way National Trail, England."}
library(png)
library(grid)
img <- readPNG("/Users/tom/Pictures/ESS-agreements.png")
grid.raster(img)
```

## The Role of National Trails

- National Trails aim to provide rewarding natural adventures  
- Quality of the scenery and landscape previously identified as primary attraction (The Countryside Agency, 2005; Wood-Gee, 2008).
- 2015 marks the 50 year anniversary of England’s first National Trail

## Measuring the effectiveness of ESS

- Measurement of ESS effectiveness has evolved
- Traditionally focused on participation levels
- Now more focused on environmental benefits delivered

## Measuring the effectiveness of ESS

- Currently no efficient method to obtain trail user opinions of ESS
- Does ESS help maintain landscape quality and character?
- Is social media data a cost-effective and efficient source?

## Data & Methods

## Twitter Data

- 60,466 geotagged tweets (generally assumed to be 1-3% of all tweets)
- Collected between 2014-06-03 and 2014-07-25 (actually 49 of the 52 days) (Lovelace, 2014)

## Pennine Way Trail

GPX file of PWNT route
used to generate the 5km and 25km trail buffers (corridors)

## ESS Data

- Shapefile of ESS agreements
- 1717 ESS agreements in place within 5km of PWNT;
- 6773 within 25km (25km used for viewshed analyses)

## Digital Elevation Model

- For viewshed analysis
- Shuttle Radar Thematic Mapper (SRTM) data (Pope, 2009) 
- 90m resolution DEM

## Land Cover Data

- The 2007 Land Cover Map (Morton et al., 2011)
- Raster 25m resolution

##  Method Overview

- Spatial selection - Twitter dataset was reduced to only the tweets which originated from within 5km of the PWNT
- Lexical Selection - relevant tweets selected using NLP; 161 remaining
- Data cleaning - The text of each tweet was processed and ‘cleaned' of spurious characters and Emoji replaced with text equivalent
- Sentiment Analysis (SentiStrength)
- Viewshed Analysis (of each overall positive & overall negative tweet)

## Sentiment Analysis

SentiStrength was used as the sentiment analysis tool (Thelwall et al., 2010) [SentiStrength Homepage](http://sentistrength.wlv.ac.uk/)

- Specifically designed for data from the social web
- Can analyse both polarity and strength of sentiment within text – although not included here
- The sum of the positive sentiment score and negative sentiment score provided the overall sentiment score for each tweet.

## Viewshed Analyses

Viewshed analyses were conducted for each overall positive and overall negative tweet:
- Viewshed Area
- Land Cover
- Ruggedness (St. dev. of elevation) 
- Extent of ESS agreements

## Selected Findings

[Link to interactive web map of PWNT user sentiment](http://tom-wilson.info.s3-website-us-west-2.amazonaws.com/PennineTweets.html)

- 40 overall positive tweets; 7 from within ESS
- 13 overall negative tweets; 2 from within ESS
- 108 overall neutral (non-sentiment bearing) tweets; 25 from within ESS

Conclusion - Sample too small to assess the effectiveness of ESS from the trail user perspective.

## Tweet Viewshed Analyses - Summary

Viewshed Analysis               |  Positive Tweets (mean)  | Negative Tweets (mean)  |
--------------------------------|--------------------------|-------------------------|
Visibility from tweet location  |           2.19           |          1.73           |
Majority land cover class       |             4            |            4            |  
Majority land cover %           |           42.13          |          36.37          |
Ruggedness Index                |           1.47           |          1.70           |
% of ESS agreements             |           65.22          |          55.66          |  

## Summary

- Findings not conclusive but a process of extracting trail user sentiment has been proposed
- Need to collect data over longer period of time
- Use hashtag and run awareness campaign?

## Known Issues

- Twitter API accessibility
- Data representation of social media data
- Danger of digital divide
- Mobile phone coverage not considered; PWNT is remote, spatial bias
- No consideration of the specific ESS options adopted by farmers and landowners.

## Further Research

- 108 tweets are not sentiment-bearing – classed as neutral
- 94 (87.03%) contained a URL within the TweetText
- Raw data - 16.18% contained a URL
- Sharing URLs has been identified as a significant aspect of Twitter use (boyd et al., 2010)
- Neutral tweets are objective with sentiment-bearing image attached?
- What sentiment is missing from these tweets?
- What do the images ‘say’?






